Array

### 2014-02-18 DOCUMENTATION/AdvancedTypes.tex
Created small introduction about shift/interval of array. Wrote about initialisation of array …

  \begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
%  	\multicolumn{3}{|c|}{Data record : \B{Record}} \\
%        \hline
%	\B{id} & : &  \B{Name}\\
%        \B{elements} & : & seq \B{Name} \\
%	\B{type} & : & \B{Struct}\\
%	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indices} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

Some examples of array are:
\begin{itemize}
	\item $array1 \in (0..10) \to INT $	(a simple array)
	\item $array2 \in (0..10) \times (0..10) \to INT $   (a multidimensional array)
	\item $array3 \in BOOL \times (0..10) \to INT$ (an array with two index types)
	\item $\{1\vdash2\vdash3, 2\vdash3\vdash4\}  \in (INT \times INT) \times INT$ (an explicit array)
\end{itemize}

\paragraph{Attributions between array and structure}

The B0 language supports attribution directly between arrays or structure, even though the attribution
must have the same domain type and range type.

@@

\subsubsection{Defining an array}

The essential elements of array are its name, size and type. The
following code declares an array global called ``Array3'' with 6000 (10*20*30)
integer numbers with 32 bits and initialized with zero.

\begin{llvmcode}
@Array3 = internal global [10 x [20 x [30 x i32]]] zeroinitializer, align 16
\end{llvmcode}

Some type examples are:
\begin{itemize}
	\item \verb|[25 x i32]| - Array with 25 elements of 32-bit integer values.
	\item \verb|[4 x [5 x [6 x i16]]] | - Multidimensional array with (4 x 5
x 6) elements of 16-bit integer values.
\end{itemize}

%TODO: Explain the other rules of Array 
%- Declaring the type of array  a : (INT X INT) --> INT
%- Declaring a array dynamically {1|->2|->3, 2|->3|->4}

\subsubsection{Getting a element from array}
\VGM{Maybe it can be summarized with text of section 3}

@@

\begin{llvmcode}
  %7 = getelementptr [10 x [20 x [30 x i32]]]* @Array3, i32 0, i64 2, i64 3, i64 4
\end{llvmcode}

The first parameter defines the type and name of the variable, and the others are
indexes to select the element in the data type. The second parameter (i32 0)
selects the first element in the pointer (* @Array3). The last elements 
(i64 2, i64 3, i64 4) selects the value from position (2,3,4) storing it in \%7.
This array is multidimensional then several index must be declared separated
by comma to get the specific element.

\iffalse
\begin{verbatim}
<temporary variable> = getelementptr  
  [<number of elements> x <element type>]   * @<variable_name>, 
  <type and index of pointer> <index_value>,
  <type and index. ex:i64>  <index_value>
\end{verbatim}
\fi

@@

\begin{ccode} 
  int x=2; int y=3; int z=4;
  Array3[x][y][z]=7;
\end{ccode}

@@

\begin{llvmcode}
%x = alloca i32, align 4
%y = alloca i32, align 4
%z = alloca i32, align 4
store i32 2, i32* %x, align 4
store i32 3, i32* %y, align 4
store i32 4, i32* %z, align 4
%1 = load i32* %z, align 4
%2 = sext i32 %1 to i64
%3 = load i32* %y, align 4
%4 = sext i32 %3 to i64
%5 = load i32* %x, align 4
%6 = sext i32 %5 to i64
%7 = getelementptr[10x[20x[30xi32]]]*@Array3,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7, align 4;setting an element
\end{llvmcode}

@@

@@

\paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indices that 
calculates the address of element. This rule also supports to translate 
expressions of arrays.

\begin{align*}
\lefteqn{\trad{n}{ArrayItem} \isdef} \\
&  \LET p_1 \sep v_1 \sep t_1  = \trad{n.base}{Exp}  \IN \\
&  \LET r \sep p_2  = \trad{n.indices}{LRExp}  \IN \\
&  \LET v =  \variable \AND  t = \trad{\B{n.type}}{Type}  \IN \\
&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0 \PH{r}\nl"}  \sep v \sep t\\
\end{align*}

A list of type and variable is generated by the following rule.
%TODO: check this rule with David

\begin{align*}
&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
 \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
\end{align*}

@@

>>>>

  \begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
%  	\multicolumn{3}{|c|}{Data record : \B{Record}} \\
%        \hline
%	\B{id} & : &  \B{Name}\\
%        \B{elements} & : & seq \B{Name} \\
%	\B{type} & : & \B{Struct}\\
%	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

Some examples of array are:
\begin{itemize}
	\item $array1 \in (0..10) \to INT $	(a simple array)
	\item $array2 \in (1..10) \times (1..20) \times (1..30) \to INT $   (a multidimensional array)
	\item $\{1\vdash1\vdash4, 1\vdash2\vdash5\}  \in (INT \times INT) \times INT$ (an initialised array)
\end{itemize}

@@

\subsubsection{Declaring an array}

The essential elements of the array are its name, size and type. 
The size of ``e'' array with the indexes (i1, ... , in) are defined according 
to the formula: $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - 
e.in\_start+1)$.

The following code declares a multidimensional global array called ``Array2'' 
with 6000 (10*20*30) integer numbers with 32 bits and initialized with zero.

\begin{llvmcode}
@Array2 = internal global [10 x [20 x [30 x i32]]] zeroinitializer, align 16
\end{llvmcode}

\VGM{It is not implemented yet, because it requires generating an XML example.}
The following example initialises a new array in C:

\begin{ccode}
int32_t arr[1][2][1]= {{{4},{5}}};
\end{ccode}

The syntax rule used to initialize an array in LLVM is:\\
$@<var\_name> = constant [<size> x <type>] [<type> <value>]$.\\
The translation of the last example is:

\begin{llvmcode}
@arr=constant[1 x [2 x [1 x i32]]] [[2 x [1 x i32]] [[1 x i32] [i32 4],[1 x i32] [i32 5]]]
\end{llvmcode}

\subsubsection{Getting a element from array}

@@

\begin{llvmcode}
%7 = getelementptr [10 x [20 x [30 x i32]]]* @Array2, i32 0, i64 2, i64 3, i64 4
\end{llvmcode}

The first parameter defines the type and name of the variable, and the others are
indexes to select the element in the data type. The second parameter (i32 0)
selects the first element in the pointer (* @Array2). The last elements 
(i64 2, i64 3, i64 4) select the value from position (2,3,4) storing it in \%7.
This array is multidimensional, so several indexes must be declared separated
by comma to get the specific element.

@@

\begin{ccode} 
int x=2; int y=3; int z=4;
Array2[x][y][z]=7;
\end{ccode}

@@

\begin{llvmcode}
%x = alloca i32, align 4
%y = alloca i32, align 4
%z = alloca i32, align 4
store i32 2, i32* %x, align 4
store i32 3, i32* %y, align 4
store i32 4, i32* %z, align 4
%1 = load i32* %z, align 4
%2 = sext i32 %1 to i64
%3 = load i32* %y, align 4
%4 = sext i32 %3 to i64
%5 = load i32* %x, align 4
%6 = sext i32 %5 to i64
%7 = getelementptr[10x[20x[30xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7, align 4;setting an element
\end{llvmcode}

@@

The element access of array can need a shift in indexes. When the domain of 
array is an interval and it is not started from zero, it must adjust the 
indexes shifting the access, because the addressing  of array in intermediate 
representation LLVM starts at 0.

@@

\paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indexes that 
calculates the address of element. This rule also supports to translate 
expressions of arrays.

\begin{align*}
\lefteqn{\trad{n}{ArrayItem} \isdef} \\
&  \LET p_1 \sep v_1 \sep t_1  = \trad{n.base}{Exp}  \IN \\
&  \LET r \sep p_2  = \trad{n.indexes}{LRExp}  \IN \\
&  \LET v =  \variable \AND  t = \trad{\B{n.type}}{Type}  \IN \\
&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0 \PH{r}\nl"}  \sep v \sep t\\
\end{align*}

A list of type and variable is generated by the following rule.
%TODO: insert a rule to do the shift address with the first element of interval
\begin{align*}
&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
 \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
\end{align*}

%TODO: support to shift between interval
\begin{itemize}
	\item $array2 \in (10..50) \times ((-10)..100) \to INT $   (a multidimensional array)
\end{itemize}

The array indexes can composed of only primitive types (INTEGER and BOOLEAN). To access an element the formula is $address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$.

%TODO create a new rule based on the below rule.

%it must change the declaration and the rule of ArrayItem

@@

The B0 language supports attribution directly between arrays or structures, even though the attribution
must have the same domain type and range type.
